Species turnover reveals hidden effects of decreasing Nitrogen deposition in mountain hay meadows

### 2018-12-13 Manuscript.Rmd
Credibility intervals mit compatibility interval ersetzt.

affiliation:
  - code: 1
    address: Zoological Institute, University of Basel, Basel, Switzerland
  - code: 2
    address: Hintermann  Weber AG, Austrasse 2a, 4153 Reinach, Switzerland
  - code: 3
    address: Meteotest, Fabrikstrasse 14, 3012 Bern, Switzerland
  - code: 4
    address: Agroscope, Swiss Soil Monitoring Network NABO, 8046 Zurich, Switzerland
  - code: 5
    address: Federal Office for the Environment (FOEN), Air Pollution Control and Chemicals Division, 3003 Bern, Switzerland
abstract: >
  Nitrogen (N) deposition is a major threat to biodiversity in many habitats. The recent introduction of cleaner technologies in Switzerland has led to a reduction in the emissions of nitrogen oxides, with a consequent decrease in N deposition. We examined different drivers of plant community change, i.e. N deposition, climate warming, and land-use change, in Swiss mountain hay meadows, using data from the Swiss biodiversity monitoring program. We compared indicator values of species that disappeared from or colonized a site (species turnover) with the indicator values of randomly chosen species from the same site. While oligotrophic plant species were more likely to colonize, compared to random expectation, we found only weak shifts in plant community composition. In particular, the average nutrient value of plant communities remained stable over time (2003-2017). We found the largest deviations from random expectation in the nutrient values of colonizing species, suggesting that N deposition or other factors that change the nutrient content of soils were important drivers of the species composition change over the last 15 years in Swiss mountain hay meadows. In addition, we observed an overall replacement of species with lower indicator values for temperature with species with higher values. Apparently, the community effects of the replacement of eutrophic species with oligotrophic species was outweighed by climate warming. Our results add to the increasing evidence that measures of community composition such as species richness or average indicator values may be relatively stable, but that this apparent stability is often accompanied by a marked turnover of species.

@@

@@

```{r communitytrendstab}
options(knitr.kable.NA = '')
kable(res, align = c("l", rep("r", 7)), digits = 2,  booktabs = T, col.names = c("Measures", "Period 1", "Period 2", "Period 3", "Trend", "5%", "95%", "Prob. for trend"), caption = "Average measures of the biodiversity endpoints for the three sampling periods (period 1: 2003-2007; period 2: 2008-2012; period: 2013-2017). The temporal trends are given as change per 10 years and were estimated from linear mixed models with normal distribution (except for species richness with Poisson distribution and a log-link function). The measure of precision for the temporal trend is given as the 5\\% and 95\\% quantiles of the marginal posterior distribution of the linear trend (90\\% credible intervals). The column 'Prob. for trend' gives the probability that the linear trend is $> 0$. Indicator values according to Landolt et al. (2010).")
```

@@

```{r difffromrandom}
getres <- function(param, gradient, file) {
  tt <- tibble()
  load(paste0("Communitysim/", file))
  tres$StandEf <- tres$meanreal - tres$mean
  mod <- stan_glm(StandEf ~ Sitemeasure, data = tres , family = gaussian(), cores = 4)
  tt[1, "Ellenberg value"] <- param
  tt[1, "Gradient"] <- gradient
  tt[1, "Estimate"] <- coef(mod)[1]
  tmp <- posterior_interval(mod, pars = "(Intercept)")[1,] %>% as.vector()
  tt[1, "5%"] <- tmp[1] 
  tt[1, "90%"] <- tmp[2] 
  tt[1, "Estimate_Slope"] <- coef(mod)[2]
  tmp <- posterior_interval(mod, pars = "Sitemeasure")[1,] %>% as.vector()
  tt[1, "5%_Slope"] <- tmp[1] 
  tt[1, "90%_Slope"] <- tmp[2] 
  tt
}

@@

```{r difffromrandomtab}
kable(res, align = c("l", "l", rep("c", 6)), digits = 3,  booktabs = T, col.names = c("Indicator value", "Gradient", "Estimate", "5%", "90%", "Estimate", "5%", "90%"), caption = "Difference in the average indicator value of species that (a) disappeared from a site or (b) newly colonized a site compared to the same number of species that were randomly selected from all species recorded at a site. Shown are the results from linear models, with the difference between disappeared/colonized species and random species as dependent variable and the site-measure (gradient) as predictor variable.") %>%  
  kable_styling() %>% 
  group_rows("(a) Plants that disappeared from a site", 1, 4, bold = F, italic = T) %>%
  group_rows("(b) Plants that newly colonized a site", 5, 8, bold = F, italic = T) %>%
  add_header_above(c(" " = 2, "Difference from random" = 3, "Change along gradient" = 3))
```

@@

```{r cl, fig.cap="Colonization (a) and local survival (b) of oligotrophic species (indicator value for nutrients $<3$; red line) and of eutrophic species (indicator value for nutrients $>3$) along the N deposition gradient. Given are means and 95\\%-credible intervals from logistic linear mixed models. The vertical lines indicate the deposition rate, with equal colonization or survival probabilities for oligotrophic and eutrophic species, the solid line indicating the median, and the dashed lines the 5\\% and 95\\% quantiles of the marginal posterior distribution."}
X <- 1:30
Xsd <- (X - 10) / 10

@@

compileres <- function(EL) {
  tt <- map(1:length(X), function(s) map(post, paste(EL)) %>% map_dbl(s))
  tibble(mean = map_dbl(tt, mean),
         low = map_dbl(tt, quantile, probs = 0.025),
         high = map_dbl(tt, quantile, probs = 0.975),
         X = X,
         value = paste(EL)
  )
}

>>>>

affiliation:
  - code: 1
    address: Zoological Institute, University of Basel, Basel, Switzerland
  - code: 2
    address: Hintermann  Weber AG, Austrasse 2a, 4153 Reinach, Switzerland
  - code: 3
    address: Meteotest, Fabrikstrasse 14, 3012 Bern, Switzerland
  - code: 4
    address: Agroscope, Swiss Soil Monitoring Network NABO, 8046 Zurich, Switzerland
  - code: 5
    address: Federal Office for the Environment (FOEN), Air Pollution Control and Chemicals Division, 3003 Bern, Switzerland
abstract: >
  Nitrogen (N) deposition is a major threat to biodiversity in many habitats. The recent introduction of cleaner technologies in Switzerland has led to a reduction in the emissions of nitrogen oxides, with a consequent decrease in N deposition. We examined different drivers of plant community change, i.e. N deposition, climate warming, and land-use change, in Swiss mountain hay meadows, using data from the Swiss biodiversity monitoring program. We compared indicator values of species that disappeared from or colonized a site (species turnover) with the indicator values of randomly chosen species from the same site. While oligotrophic plant species were more likely to colonize, compared to random expectation, we found only weak shifts in plant community composition. In particular, the average nutrient value of plant communities remained stable over time (2003-2017). We found the largest deviations from random expectation in the nutrient values of colonizing species, suggesting that N deposition or other factors that change the nutrient content of soils were important drivers of the species composition change over the last 15 years in Swiss mountain hay meadows. In addition, we observed an overall replacement of species with lower indicator values for temperature with species with higher values. Apparently, the community effects of the replacement of eutrophic species with oligotrophic species was outweighed by climate warming. Our results add to the increasing evidence that plant communities in changing environments may be relatively stable regarding average community composition, but that this apparent stability is often accompanied by a marked turnover of species..

@@

As a summary of the result of a parameter we give the 5% and 95% quantiles of the marginal posterior distribution, which we interpreted as a compatibility interval showing effect sizes most compatible with the data, under the model and prior distribution used to compute the interval [@Amrhein2018].

@@

```{r communitytrendstab}
options(knitr.kable.NA = '')
kable(res, align = c("l", rep("r", 7)), digits = 2,  booktabs = T, col.names = c("Measures", "Period 1", "Period 2", "Period 3", "Trend", "5%", "95%", "Prob. for trend"), caption = "Average measures of the biodiversity endpoints for the three sampling periods (period 1: 2003-2007; period 2: 2008-2012; period: 2013-2017). The temporal trends are given as change per 10 years and were estimated from linear mixed models with normal distribution (except for species richness with Poisson distribution and a log-link function). The measure of precision for the temporal trend is given as the 5\\% and 95\\% quantiles of the marginal posterior distribution of the linear trend (90\\% compatibility interval). The column 'Prob. for trend' gives the probability that the linear trend is $> 0$. Indicator values according to Landolt et al. (2010).")
```

@@

```{r difffromrandom}
getres <- function(param, gradient, file) {
  tt <- tibble()
  load(paste0("Communitysim/", file))
  tres$StandEf <- tres$meanreal - tres$mean
  mod <- stan_glm(StandEf ~ Sitemeasure, data = tres , family = gaussian(), cores = 4)
  tt[1, "Ellenberg value"] <- param
  tt[1, "Gradient"] <- gradient
  tt[1, "Estimate"] <- coef(mod)[1]
  tmp <- posterior_interval(mod, pars = "(Intercept)")[1,] %>% as.vector()
  tt[1, "5%"] <- tmp[1] 
  tt[1, "95%"] <- tmp[2] 
  tt[1, "Estimate_Slope"] <- coef(mod)[2]
  tmp <- posterior_interval(mod, pars = "Sitemeasure")[1,] %>% as.vector()
  tt[1, "5%_Slope"] <- tmp[1] 
  tt[1, "95%_Slope"] <- tmp[2] 
  tt
}

@@

```{r difffromrandomtab}
kable(res, align = c("l", "l", rep("c", 6)), digits = 3,  booktabs = T, col.names = c("Indicator value", "Gradient", "Estimate", "5%", "95%", "Estimate", "5%", "95%"), caption = "Difference in the average indicator value of species that (a) disappeared from a site or (b) newly colonized a site compared to the same number of species that were randomly selected from all species recorded at a site. Shown are the results from linear models, with the difference between disappeared/colonized species and random species as dependent variable and the site-measure (gradient) as predictor variable.") %>%  
  kable_styling() %>% 
  group_rows("(a) Plants that disappeared from a site", 1, 4, bold = F, italic = T) %>%
  group_rows("(b) Plants that newly colonized a site", 5, 8, bold = F, italic = T) %>%
  add_header_above(c(" " = 2, "Difference from random" = 3, "Change along gradient" = 3))
```

@@

```{r cl, fig.cap="Colonization (a) and local survival (b) of oligotrophic species (indicator value for nutrients $<3$; red line) and of eutrophic species (indicator value for nutrients $>3$) along the N deposition gradient. Given are means and 95\\% compatibility intervals from logistic linear mixed models. The vertical lines indicate the deposition rate, with equal colonization or survival probabilities for oligotrophic and eutrophic species, the solid line indicating the median, and the dashed lines the 5\\% and 95\\% quantiles of the marginal posterior distribution."}
X <- 1:30
Xsd <- (X - 10) / 10

@@

compileres <- function(EL) {
  tt <- map(1:length(X), function(s) map(post, paste(EL)) %>% map_dbl(s))
  tibble(mean = map_dbl(tt, mean),
         low = map_dbl(tt, quantile, probs = 0.05),
         high = map_dbl(tt, quantile, probs = 0.95),
         X = X,
         value = paste(EL)
  )
}
